Community Embeddings with Bayesian Gaussian Mixture Model and Variational Inference\\ \thanks{The reported study was partially supported by RFBR grant \textnumero 20-07-00958. The paper was prepared within the framework of the HSE University Project Group Competition 2020-2022.}

### 2020-10-14 paper-ComE-BGMM-VI.tex
write conclusion

In their original \citeyear{ComE} paper \textit{Learning Community Embedding with Community Detection and Node Embedding on Graphs}, \citeauthor{ComE} identify three evaluation tasks for experimentation: graph visualization, community detection and node classification \cite{ComE}. Community detection and node classification can objectivelly be quantitatively evaluated, while graph visualization is subjectively evaluated.

@@

\begin{figure}[H]
    \centering
    \begin{subfigure}{.5\linewidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.02054029917392275	& 0.021070234113712373	\\
                5	& 0.030399077942659558	& 0.02551070276610302	\\
                10	& 0.01722986509950581	& 0.030197444831591175					\\
                20	& 0.029249335242380856	& 0.03289726818006926	\\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    point meta={y*100},
                    yticklabel= {\pgfmathparse{\tick*100}\pgfmathprintnumber{\pgfmathresult}\%},
                    ymin=0, ymax=0.15,
                    xlabel={$K$},
                    ylabel={Conductance},
                    nodes near coords={\pgfmathprintnumber\pgfplotspointmeta\%},
                    nodes near coords align={vertical},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{two-dimensional embeddings.}
        \label{fig:dblp_conductance_d2}
    \end{subfigure}%
    %
    \begin{subfigure}{.5\linewidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.02072897576868255	& 0.05153027478804642	\\
                5	& 0.013701852396354014	& 0.011492883939736796	\\
                10	& 0.06247833021288399	& 0.04919872794163497					\\
                20	& 0.11969825263644061	& 0.07018888966334426 \\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    point meta={y*100},
                    yticklabel= {\pgfmathparse{\tick*100}\pgfmathprintnumber{\pgfmathresult}\%},
                    ymin=0, ymax=0.15,
                    xlabel={$K$},
                    nodes near coords={\pgfmathprintnumber\pgfplotspointmeta\%},
                    nodes near coords align={vertical},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{128-dimensional embeddings.}
        \label{fig:dblp_conductance_d128}
    \end{subfigure}

    \caption{Conductance on DBLP dataset. Lower conductance is better.}
    \label{fig:dblp_conductance}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.4636647057169119	& 0.4623820822063423	\\
                5	& 0.46257130393293455		& 0.46502447614120096		\\
                10	& 0.43349289423082416		& 0.46005304584573076		\\
                20	& 0.3999601374635516	& 0.45371067853180896	\\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    nodes near coords,
                    nodes near coords align={vertical},
                    ymin=0,ymax=1,
                    xlabel={$K$},
                    ylabel={NMI},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{two-dimensional embeddings.}
        \label{fig:dblp_nmi_d2}
    \end{subfigure}%
    %
    \begin{subfigure}{.5\textwidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.4802479401582417	& 0.43575239496704926	\\
                5	& 0.6746848015894669		& 0.668553614850714		\\
                10	& 0.4963225850192131		& 0.47369605057376785		\\
                20	& 0.3018911505699723	& 0.30816578800908423	\\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    nodes near coords,
                    nodes near coords align={vertical},
                    ymin=0,ymax=1,
                    xlabel={$K$},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{128-dimensional embeddings.}
        \label{fig:dblp_nmi_d128}
    \end{subfigure}

    \caption{NMI on DBLP dataset. Higher NMI is better.}
    \label{fig:dblp_nmi}
\end{figure}

@@

Figure \ref{table:dblp_f1} shows the results of ComE GMM+EM and ComE BGMM+VI on the task node classification compared over multiple values for the hyperparameter number of communities $K$. The evaluation metrics micro- and macro-F1 used for node classification are described in Section \ref{sec:come_eval_metrics_nc}.

@@

For NMI, ComE BGMM+VI seems to perform slightly better than ComE when choosing a two-dimensional embedding instead of a 128-dimensional embedding and larger initial number of communities that the true number of communities: $K>K^*$. It is hypothesized that this advantage comes from the ability of BGMM+VI to drop unused components in lower-dimensional space. It appears that BGMM+VI does not drop components as easy in higher-dimensional space.

Examining the weights of components produced by ComE BGMM+VI in two- and 128-dimensional space provides the insight, that ComE BGMM+VI produces the same distributions of 5 components' weights for $K=5$, $K=10$, and $K=20$ in two-dimensional space. In 128-dimensional space, on the other hand, ComE BGMM+VI does not drop components and remains with weight distributions including all the components added at initialization. In future work, it might be interesting to explore the weight concentration prior or other mechanisms of the BGMM+VI to allow ComE to drop unused communities at higher dimensionality as well.

Evaluation of node classification aims to quantify the quality of classifications generated by node embeddings. Firstly, embeddings are generated for the whole graph. Then, a support vector machine (SVM) is trained with $90\%$ of the 128-dimensional node embeddings as features and true labels as target values. Ultimately, the micro- and macro-F1 scores are determined for the remaining $10\%$ testing data. In the experiment of the DBLP dataset, no advantage of using ComE BGMM+VI over GMM+EM was identified for node classification. It is reasonable to note, that node classification with micro- and macro-F1 of over $90\%$ is achieved, with the 128-dimensional embedding, leaving little room for optimization.

@@

>>>>

In their original \citeyear{ComE} paper \textit{Learning Community Embedding with Community Detection and Node Embedding on Graphs}, \citeauthor{ComE} identify three evaluation tasks for experimentation: graph visualization, community detection and node classification \cite{ComE}. Community detection and node classification can objectively be quantitatively evaluated, while graph visualization is subjectively evaluated.

@@

\begin{figure*}[htbp]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.02054029917392275	& 0.021070234113712373	\\
                5	& 0.030399077942659558	& 0.02551070276610302	\\
                10	& 0.01722986509950581	& 0.030197444831591175					\\
                20	& 0.029249335242380856	& 0.03289726818006926	\\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=.5\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    point meta={y*100},
                    yticklabel= {\pgfmathparse{\tick*100}\pgfmathprintnumber{\pgfmathresult}\%},
                    ymin=0, ymax=0.15,
                    xlabel={$K$},
                    ylabel={Conductance},
                    nodes near coords={\pgfmathprintnumber\pgfplotspointmeta\%},
                    nodes near coords align={vertical},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{two-dimensional embeddings.}
        \label{fig:dblp_conductance_d2}
    \end{subfigure}%
    %
    \begin{subfigure}{.5\textwidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.02072897576868255	& 0.05153027478804642	\\
                5	& 0.013701852396354014	& 0.011492883939736796	\\
                10	& 0.06247833021288399	& 0.04919872794163497					\\
                20	& 0.11969825263644061	& 0.07018888966334426 \\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=.5\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    point meta={y*100},
                    yticklabel= {\pgfmathparse{\tick*100}\pgfmathprintnumber{\pgfmathresult}\%},
                    ymin=0, ymax=0.15,
                    xlabel={$K$},
                    nodes near coords={\pgfmathprintnumber\pgfplotspointmeta\%},
                    nodes near coords align={vertical},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{128-dimensional embeddings.}
        \label{fig:dblp_conductance_d128}
    \end{subfigure}

    \caption{Conductance on DBLP dataset. Lower conductance is better.}
    \label{fig:dblp_conductance}
\end{figure*}

@@

\begin{figure*}[htbp]
    \centering
    \begin{subfigure}{.5\textwidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.4636647057169119	& 0.4623820822063423	\\
                5	& 0.46257130393293455		& 0.46502447614120096		\\
                10	& 0.43349289423082416		& 0.46005304584573076		\\
                20	& 0.3999601374635516	& 0.45371067853180896	\\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=.5\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    nodes near coords,
                    nodes near coords align={vertical},
                    ymin=0,ymax=1,
                    xlabel={$K$},
                    ylabel={NMI},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{two-dimensional embeddings.}
        \label{fig:dblp_nmi_d2}
    \end{subfigure}%
    %
    \begin{subfigure}{.5\textwidth}
        \centering
        \begin{tikzpicture}
            \pgfplotstableread[row sep=\\,col sep=&]{
                K	& GMM					& BGMM					\\
                2	& 0.4802479401582417	& 0.43575239496704926	\\
                5	& 0.6746848015894669		& 0.668553614850714		\\
                10	& 0.4963225850192131		& 0.47369605057376785		\\
                20	& 0.3018911505699723	& 0.30816578800908423	\\
            }\mydata
            \begin{axis}[
                    ybar,
                    bar width=.5cm,
                    width=\linewidth,
                    height=.5\linewidth,
                    legend style={at={(0.5,1)},
                            anchor=north,legend columns=-1},
                    symbolic x coords={2,5,10,20},
                    xtick=data,
                    nodes near coords,
                    nodes near coords align={vertical},
                    ymin=0,ymax=1,
                    xlabel={$K$},
                ]
                \addplot table[x=K,y=GMM]{\mydata};
                \addplot table[x=K,y=BGMM]{\mydata};
                \legend{GMM, BGMM}
            \end{axis}
        \end{tikzpicture}
        \caption{128-dimensional embeddings.}
        \label{fig:dblp_nmi_d128}
    \end{subfigure}

    \caption{NMI on DBLP dataset. Higher NMI is better.}
    \label{fig:dblp_nmi}
\end{figure*}

@@

Figure \ref{table:dblp_f1} shows the results of ComE GMM+EM and ComE BGMM+VI on the task node classification compared over multiple values for the hyperparameter number of communities $K$. The evaluation metrics micro- and macro-F1 are used.

@@

For NMI, ComE BGMM+VI seems to perform slightly better than ComE when choosing a two-dimensional embedding instead of a 128-dimensional embedding and a greater initial number of communities that the true number of communities: $K>5$. It is hypothesized that this advantage comes from the ability of BGMM+VI to drop unused components in lower-dimensional space. It appears that BGMM+VI does not drop components as easy in higher-dimensional space.

Examining the weights of components produced by ComE BGMM+VI in two- and 128-dimensional space provides the insight, that ComE BGMM+VI produces the same distributions of 5 components' weights for $K=5$, $K=10$, and $K=20$ in two-dimensional space. In 128-dimensional space, on the other hand, ComE BGMM+VI does not drop components and remains with weight distributions including all the components added at initialization.

In the experiment on the DBLP dataset, no advantage of using ComE BGMM+VI over GMM+EM was identified for node classification by SVM. It is reasonable to note, that node classification with micro- and macro-F1 of over $90\%$ is achieved, with the 128-dimensional embedding, leaving little room for optimization.

@@

\section{Conclusion}

ComE BGMM+VI, an extension of the community embedding algorithm ComE is presented in this paper. ComE BGMM+VI utilizes a Bayesian Gaussian mixture model instead of a non-Bayesian Gaussian mixture model for community embedding and detection. The Bayesian community embedding algorithm ComE BGMM+VI is compared to ComE GMM+EM visually on the Karate Club graph dataset and quantitatively evaluated on community detection and node classification the DBLP dataset.

Visually comparing ComE BGMM+VI to ComE GMM+EM clearly shows the advantage of ComE BGMM+VI over ComE GMM+EM for an unknown number of communities $K$ and a low representation size $D$. ComE BGMM+VI is able to drop unused communities and reduce $K$ to the number of communities present at the resolution captured at the chosen representation size $D$ through the weight concentration prior hyperparameter trade-off parameter.

A slight edge of ComE BGMM+VI over the original ComE as identified for community detection NMI in a low-dimensional embedding for $K$ larger than the true number of communities. More notably, ComE BGMM+VI captures 5 communities in the DBLP dataset at representation size $D=2$ for all tested values for the number of communities $K>5$. The true number of communities in the DBLP dataset is 5. ComE with non-Bayesian Gaussian mixture model is not able to drop communities for $K>5$, as seen in Figure \ref{fig:dblp_weights}.
